---
title: Exploratory Data Analysis - White Wine Quality
author: Julia Fuhrman
output:
  html_document: default
  pdf_document: default
---
================================================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(ggthemes)
library(gridExtra)
library(GGally)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
setwd('/Users/julia/Documents/UdacityProjects/Data Analyst Nanodegree/Exploratory Data Analysis/final-project')
WhiteWineData <- read.csv('wineQualityWhites.csv')
theme_set(theme_minimal(12)) 
```

This analysis explores 11 chemical properties of white wine for 4,898 wine 
samples, as well as a rating of the quality of the wine on a 0-10 scale.

# Univariate Plots Section

### 1. Fixed Acidity (tartaric acid)

"Fixed Acidity" was measured by the amount of tartaric acid in the wine in grams
per cubic decimeter.  This acid decreases as grapes mature on the vine, so the 
riper the wine, the lower the amount of tartaric acid.  This acid contributes to
the "tartness" of the wine.  The distribution of this variable in the dataset is
normal, with a mean of 6.855 g/dm^3 and a median of 6.8 g/dm^3.

```{r echo=FALSE, Fixed_Acidity_Summary}
summary(WhiteWineData$fixed.acidity)
```

```{r echo=FALSE, Fixed_Acidity}
qplot(data = WhiteWineData,
      x = WhiteWineData$fixed.acidity, 
      binwidth = .1,
      xlab = "g / dm^3",
      color = I("black"),
      fill = I("pink")) +
  scale_x_continuous(lim = c(3,15), 
                     breaks = seq(3,15,1))

```

### 2. Volatile Acidity (acetic acid)

"Volatile Acidity" was measured by the amount of acetic acid in the wine in 
grams per cubic decimeter.  The normal amount found in wine is around 
0.3 g/dm^3.  This acid contributes to the smell and taste of the wine.  As the 
amount of acetic acid increases above 0.3 g/dm^3, the wine develops a sour, 
vinegar taste to it.  The overall distribution of this variable is slightly positively skewed.
The range within the 1st quartile was 0.13 g/dm^3, while the range within the 
4th quartile was 0.78 g/dm^3 with 50% of the data falling between 0.27 g/dm^3 
and 0.39 g/dm^3

```{r echo=FALSE, Volatile_Acidity_Summary}
summary(WhiteWineData$volatile.acidity)
```

```{r echo=FALSE, Volatile_Acidity}
qplot(data = WhiteWineData, 
      x = WhiteWineData$volatile.acidity, 
          binwidth = .01,
          xlab = "g / dm^3",
          color = I("black"),
          fill = I("red")) +
  scale_x_continuous(lim = c(0,1.5), 
                     breaks = seq(0,1.5,.1))
```

### 3. Citric Acid

The amount of citric acid in the wine was measured in grams per cubic decimeter.
This acid is used in small quantities to lower the pH (make it slightly more
acidic) as well as add 'freshness' and flavor.  The distribution of this 
variable in the dataset is slightly positively skewed, with a mean of 
0.3342 g/dm^3 and a median of 0.3200 g/dm^3.

```{r echo=FALSE, Citric_Acid_Summary}
summary(WhiteWineData$citric.acid)
```

```{r echo=FALSE, Citric_Acid}
qplot(data = WhiteWineData, 
      x = WhiteWineData$citric.acid, 
      binwidth = .05,
      xlab = "g / dm^3",
      color = I("black"),
      fill = I("orange")) +
  scale_x_continuous(lim = c(0,2), breaks = seq(0,2,.2))
```

### 4. Residual Sugar

"Residual Sugar" is a measure of the amount of sugar remaining after 
fermentation stops in grams per cubic decimeter. It's rare to find wines with 
less than 1 g/dm^3 and wines with greater than 45 g/dm^3 are considered sweet. 
The outliers in this dataset cause the distribution to be non-normal with the 
range within the 4th quartile of 55.9 g/dm^3 and 75% of the data falling below 
9.9 g/dm^3.  After transforming this data to a log10 scale, it appears the data 
is bimodal. 

```{r echo=FALSE, Residual_Sugar_Summary}
summary(WhiteWineData$residual.sugar)
```

```{r echo=FALSE, Residual_Sugar}
residual.sugar1 <- qplot(data = WhiteWineData, 
                         x = WhiteWineData$residual.sugar,
                         binwidth = .1,
                         xlab = "g / dm^3",
                         color = I("black"),
                         fill = I("dark grey"))

residual.sugar2 <- residual.sugar1 + scale_x_log10()

grid.arrange(residual.sugar1, residual.sugar2)
```

### 5. Chlorides (sodium chloride)

"Chlorides" were the measure of the amount of sodium chloride (salt) in the wine
in g/dm^3.  This dataset has a positively skewed distribution with 75% of the
data falling below 0.05 g/dm^3 and 25% of the data falling between 0.05 g/dm^3 
and 0.346 g/dm^3. 

```{r echo=FALSE, Chlorides_Summary}
summary(WhiteWineData$chlorides)
```

```{r echo=FALSE, Chlorides}
qplot(data = WhiteWineData, 
      x = WhiteWineData$chlorides, 
      binwidth = .005,
      xlab = "g / dm^3",
      color = I("black"),
      fill = I("blue")) +
  scale_x_continuous(lim = c(0,.35), breaks = seq(0,.35,.05))
```

### 6. Free Sulfur Dioxide

Free Sulfur Dioxide was measured in mg/dm^3.  The free form of SO2 exists in 
equilibrium between molecular SO2 (as a dissolved gas) and bisulfite ion.  It 
prevents microbial growth and the oxidation of wine.  This variable in the 
dataset is positively skewed because of larger numbers within a range of 
243 mg/dm^3 in the 4th quartile of this dataset.  The median is 34 mg/dm^3 and 
the mean is 35.31 mg/dm^3.  

```{r echo=FALSE, Free_Sulfur_Dioxide_Summary}
summary(WhiteWineData$free.sulfur.dioxide)
```

```{r echo=FALSE, Free_Sulfur_Dioxide}
qplot(data = WhiteWineData, 
      x = WhiteWineData$free.sulfur.dioxide, 
      binwidth = 5,
      xlab = "mg / dm^3",
      color = I("black"),
      fill = I("yellow")) +
  scale_x_continuous(lim = c(0,300), breaks = seq(0,300,20))

```

### 7. Total Sulfur Dioxide

"Total Sulfur Dioxide" is the amount of free and bound forms of S02.  In low 
concentrations, SO2 is mostly undetectable in wine, but at free SO2 
concentrations over 50 ppm, SO2 becomes evident in the smell and taste of wine.
Outliers skew this distribution in the positive direction with a median of 
134 mg/dm^3 and a mean of 138.4 mg/dm^3.  50% of the data falls between 
108 mg/dm^3 and 167 mg/dm^3.

```{r echo=FALSE, Total_Sulfur_Dioxide_Summary}
summary(WhiteWineData$total.sulfur.dioxide)
```

```{r echo=FALSE, Total_Sulfur_Dioxide}
qplot(data = WhiteWineData,
      x = WhiteWineData$total.sulfur.dioxide, 
      binwidth = 10,
      xlab = "mg / dm^3",
      color = I("black"),
      fill = I("brown")) +
  scale_x_continuous(lim = c(0,450), breaks = seq(0,450,20))
```

### 8. Density

"Density" is determined by the alcohol and sugar content of the wine in grams 
per cubic centimeter.  50% of the data falls beween 0.9917 g/cm^3 and 0.9961 
g/cm^3.  The median is 0.9937 and the mean is 0.9940.  This distribution in 
slightly positively skewed because of outliers falling between 0.9961 g/cm^3 and
1.039 g/cm^3.

```{r echo=FALSE, Density_Summary}
summary(WhiteWineData$density)
```

```{r echo=FALSE, Density}
qplot(data = WhiteWineData, 
      x = WhiteWineData$density,
      binwidth = .001,
      xlab = "g / cm^3",
      color = I("black"),
      fill = I("green")) +
  scale_x_continuous(lim = c(.98,1.05), 
                     breaks = seq(.98,1.05,.01))
```

### 9. pH

"pH" describes how acidic or basic a wine is on a scale from 0 (very acidic) to 
14 (very basic); most wines are between 3-4 on the pH scale.  This variable is 
normally distributed in this dataset with a mean of 3.188 and a median of 3.180.

```{r echo=FALSE, pH_Summary}
summary(WhiteWineData$pH)
```

```{r echo=FALSE, pH}
qplot(data = WhiteWineData, 
      x = WhiteWineData$pH, 
      binwidth = .05,
      xlab = "pH",
      color = I("black"),
      fill = I("purple"))+
  scale_x_continuous(lim = c(2.5,4), 
                     breaks = seq(2.5,4,.1))
```

### 10. Sulphates (postassium sulphate)

"Sulphates" are a measurement of the amount of postassium sulphate in g/dm^3.  
Potassium sulphate is a wine additive which can contribute to sulfur dioxide gas
(S02) levels, which acts as an antimicrobial and antioxidant.  This variable is 
slightly positively skewed, with a median of 0.47 g/dm^3 and a mean of 0.4898 
g/dm^3.  50% of the data falls between 0.41 g/dm^3 and 0.55 g/dm^3.

```{r echo=FALSE, Sulphates_Summary}
summary(WhiteWineData$sulphates)
```

```{r echo=FALSE, Sulphates}
qplot(data = WhiteWineData,
      x = WhiteWineData$sulphates, 
      binwidth = .05,
      xlab = "g / dm^3",
      color = I("black"),
      fill = I("dark green"))+
  scale_x_continuous(lim = c(0,1.1),
                     breaks = seq(0,1.1,.1))

```

### 11. Alcohol

"Alcohol" is a measurement of the percentage of alcohol content by volume found 
in the wine.  This variable is positively skewed with 75% of the data falling 
between 8% and 11.4%, while the remaining 25% of data fell between 11.4% and 
14.2%.

```{r echo=FALSE, Alcohol_Summary}
summary(WhiteWineData$alcohol)
```

```{r echo=FALSE, Alcohol}
qplot(data = WhiteWineData, 
      x = WhiteWineData$alcohol, 
      binwidth = .5,
      xlab = "% by volume",
      color = I("black"),
      fill = I("light blue"))+
  scale_x_continuous(lim = c(5,15), 
                     breaks = seq(5,15,1))
```

### 12. Quality

The quality of the wine was measured as the average quality of each wine, which 
was rated from 0-10 (0 = very bad, 10 = very excellent) by at least 3 experts 
based on sensory data.  This distribution would ideally have the same amount of
observations for each categorical rating, however there are no observations of
wines with ratings below 2 or any with a perfect 10 rating.  92.6% of the wines 
in this dataset were rated between 5-7.  This is a significant disadvantage for 
being able to draw conclusions on the differences between each level of quality
since each level has such a drastic difference in sample sizes.  To help 
analysis, I decided to create a new variable called "quality.bins" where I 
seperated the quality variable into 3 groups: "bad" (ratings from 0-4), 
"average" (ratings from 5-6) and "good" (ratings from 7-10). 

```{r echo=FALSE, Quality_Summary}
summary(WhiteWineData$quality)

summary(WhiteWineData$quality.bins)
```


```{r echo=FALSE, Quality}
quality1 <- qplot(data = WhiteWineData, 
                  x = WhiteWineData$quality, 
                  xlab = "rating",
                  color = I("black"),
                  fill = I("light grey"))

quality2 <- qplot(data = WhiteWineData, 
                  x = WhiteWineData$quality.bins, 
                  xlab = "rating",
                  color = I("black"),
                  fill = I("light grey"))

grid.arrange(quality1, quality2)
```

# Univariate Analysis

### What is the structure of your dataset?

This dataset includes 4,898 observations of wine with 11 numerical input
variables of chemical properties (fixed acidity, volatile acidity, citric acid, 
residual sugar, chlorides, free sulfer dioxide, total sulfur dioxide, density, 
pH, sulphates and alcohol) and 1 output variable of quality rating on a scale of
0-10, which I transformed into a new variable "quality.bins" to group the 
quality into 3 factor levels.  

```{r echo=FALSE}
str(WhiteWineData)
```

### What is/are the main feature(s) of interest in your dataset?

The guiding question of this dataset is "Which chemical properties influence the
quality of white wines?", so the main feature of interest is the quality rating.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

I think Chlorides, Density, Alcohol, Volatile Acidity and Free Sulphur Dioxide
levels might show a relationship to the quality rating of each wine.  Fixed 
Acidity and pH might also contribute slightly to the quality ratings.

### Did you create any new variables from existing variables in the dataset?

Yes, I decided to create a new variable called "quality.bins", explained in the
plots above. Using this grouping allowed me to see bigger differences between 
good and bad wine in the box plots of each variable across quality levels.  
For example, in the boxplots below for volatile acidity, when looking at all 
levels, a clear difference between patterns does not jump out at me, however 
when grouped into 3 categories, it's clear that bad wines have higher levels of 
acetic acid on average, compared to average and good wines.

```{r echo=FALSE, Category_Example}
example1 <- qplot(x = quality, 
                  y = volatile.acidity, 
                  data = WhiteWineData, 
                  geom = 'boxplot',
                  xlab = "rating",
                  ylab = "g / dm^3") +
  coord_cartesian(ylim=c(.15,.5))

grid.arrange(example1, volatile.acidity2)
```

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

The residual sugar distribution was positively skewed by outliers, so 
tranforming the x-axis to a log10 scale uncovered a bimodal distribution among 
the majority of the data.  As well, I transformed the quality variable to an 
ordered factor so that box plots would show each individual rating as its own
category.  

# Bivariate Plots Section

```{r echo=FALSE, Output_Variables}
# Create new variable
WhiteWineData$quality.bins <- cut(WhiteWineData$quality, 
                                  breaks = c(0, 4, 6, 10), 
                                  labels = c('bad','average','good')) 
summary(WhiteWineData$quality.bins)

# Convert quality ratings to factored levels
WhiteWineData$quality <- factor(WhiteWineData$quality, 
                                levels = c(0:10), 
                                order = TRUE)
```

### Comparing Quality 
volatile acidity =   As the boxplot below confirms this, bad wines have a 
median acetic acid level slightly above .3, while average to good wines have 
less.  

chlorides =  Looking at the boxplots of Chloride distribtion across each
quality rating, it looks like good wines have a lower level of sodium chloride
on average compared to bad and average wines.  

free sulfur dioxide = The boxplot shows that bad wines have a lower free 
sulfur dioxide level compared to average and good wines.  

```{r echo=FALSE, Box_Plots}
fixed.acidity <- qplot(x = quality.bins, 
                       y = fixed.acidity, 
                       data = WhiteWineData, 
                       geom = 'boxplot',
                       xlab = "rating",
                       ylab = "g / dm^3") +
  coord_cartesian(ylim=c(6,8))

volatile.acidity <- qplot(x = quality.bins, 
                          y = volatile.acidity,
                          data = WhiteWineData, 
                          geom = 'boxplot',
                          xlab = "rating",
                          ylab = "g / dm^3") +
  coord_cartesian(ylim=c(.15,.5))

citric.acid <- qplot(x = quality.bins,
                     y = citric.acid,
                     data = WhiteWineData,
                     geom = 'boxplot',
                     xlab = "rating",
                     ylab = "g / gm^3") +
  coord_cartesian(ylim=c(.15,.45))

residual.sugar <- qplot(x = quality.bins,
                        y = residual.sugar,
                        data = WhiteWineData,
                        geom = 'boxplot',
                        xlab = "rating",
                        ylab = "g / dm^3") + 
  coord_cartesian(ylim=c(0,12.5))

chlorides <- qplot(x = quality.bins,
                   y = chlorides,
                   data = WhiteWineData,
                   geom = 'boxplot',
                   xlab = "rating",
                   ylab = "g / dm^3") +
  coord_cartesian(ylim=c(0.025,.06))

free.sulfur.dioxide <- qplot(x = quality.bins,
                             y = free.sulfur.dioxide,
                             data = WhiteWineData,
                             geom = 'boxplot',
                             xlab = "rating",
                             ylab = "mg / dm^3") +
  coord_cartesian(ylim=c(0,60))

total.sulfur.dioxide <- qplot(x = quality.bins, 
                              y = total.sulfur.dioxide, 
                              data = WhiteWineData, 
                              geom = 'boxplot',
                              xlab = "rating",
                              ylab = "mg / dm^3") +
  coord_cartesian(ylim=c(75,200))

density <- qplot(x = quality.bins, 
                  y = density, 
                  data = WhiteWineData, 
                  geom = 'boxplot',
                  xlab = "rating",
                  ylab = "g / cm^3") +
  coord_cartesian(ylim=c(.99,.9975))

pH <- qplot(x = quality.bins, 
             y = pH, 
             data = WhiteWineData, 
             geom = 'boxplot',
             xlab = "rating",
             ylab = "pH") +
  coord_cartesian(ylim=c(3,3.35))

sulphates <- qplot(x = quality.bins, 
                    y = sulphates,
                    data = WhiteWineData,
                    geom = 'boxplot',
                    xlab = "rating",
                    ylab = "g / dm^3")+
  coord_cartesian(ylim=c(.35,.6))

alcohol <- qplot(x = quality.bins, 
                  y = alcohol, 
                  data = WhiteWineData, 
                  geom = 'boxplot',
                  xlab = "rating",
                  ylab = "% by volume")+
  coord_cartesian(ylim=c(9,13))

grid.arrange(fixed.acidity, volatile.acidity, citric.acid, residual.sugar, chlorides, free.sulfur.dioxide, total.sulfur.dioxide, density, pH, sulphates, alcohol, ncol= 2)
```

To analyze pairs of variables, I decided to create a scatterplot matrix to see 
which combinations of variables had the most meaningful correlation
coefficients.  

```{r echo=FALSE, Bivariate_Plots}
bivariate_scatter_matrix <- ggpairs(subset(WhiteWineData, select = -c(X, quality, quality.bins)))

bivariate_scatter_matrix
```

### Density vs Total Sulfur Dioxide

```{r echo=FALSE, Density_vs_Total_Sulfur_Dioxide}
density_vs_total.sulfur.dioxide <- bivariate_scatter_matrix[8,7]
density_vs_total.sulfur.dioxide <- density_vs_total.sulfur.dioxide +
  coord_cartesian(xlim = c(0,300), ylim=c(0.985, 1.005)) +
  geom_smooth(method='lm', color = I('green'), linetype = 2)
density_vs_total.sulfur.dioxide 
```

### Alcohol vs Chlorides

```{r echo=FALSE, Alcohol_vs_Chlorides}
alcohol_vs_chlorides <- bivariate_scatter_matrix[11,5]
alcohol_vs_chlorides <- alcohol_vs_chlorides +
  geom_smooth(method='auto', color = I('red'), linetype = 2)
alcohol_vs_chlorides 
```


# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

### What was the strongest relationship you found?
I used the following rule of thumb to determine which pairs of
variables I would analyze further:

correlation coefficient of +- 0.3 = small correlation
correlation coefficient of +- 0.5 = moderate correlation
correlation coefficient of +- 0.7 = significant correlation

The top 3 strongest relationships I found were:

- Density and Residual Sugar = 0.839
- Density and Alcohol = -0.78
- Free Sulfur Dioxide and Total Sulfur Dioxide = 0.616

Since Density is determined by the alcohol and sugar content in the wine, it is 
not surprising that both alcohol and sugar would be highly correlated with
density.  As well, free sulfur dioxide is a component of the total sulfur
dioxide, so this correlation would be expected as well.  

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

- Density and Total Sulfur Dioxide = 0.53
- Alcohol and Chlorides = -0.36

(Alcohol and Total Sulfur Dioxide = -0.449
Sugar and Total Sulfur Dioxide = 0.401
pH and Fixed Acidity = -0.426)

# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots}
multivariate_scatter_matrix  <- ggpairs(WhiteWineData, columns = c('density', 'chlorides', 'alcohol', 'total.sulfur.dioxide'), mapping = aes(color = quality))

multivariate_scatter_matrix
```
```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!