---
title: Exploratory Data Analysis - White Wine Quality
author: Julia Fuhrman
output:
  html_document: default
  pdf_document: default
---
================================================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(ggthemes)
library(gridExtra)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
setwd('/Users/julia/Documents/UdacityProjects/Data Analyst Nanodegree/Exploratory Data Analysis/final-project')
WhiteWineData <- read.csv('wineQualityWhites.csv')
theme_set(theme_minimal(12)) 
```

This analysis explores 11 chemical properties of white wine for 4,898 wine samples, as well as a rating of the quality of the wine on a 0-10 scale.

# Univariate Plots Section

```{r echo=FALSE, Univariate_Plots}
WhiteWineData$quality <- factor(WhiteWineData$quality, levels = c(0:10), order = TRUE)
```

### 1. Fixed Acidity (tartaric acid)

"Fixed Acidity" was measured by the amount of tartaric acid in the wine in grams per cubic decimeter.  This acid decreases as grapes mature on the vine, so the riper the wine, the lower the amount of tartaric acid.  This acid contributes to the "tartness" of the wine.  The distribution of this variable in the dataset is normal, with a mean of 6.855 g/dm^3 and a median of 6.8 g/dm^3.   

```{r echo=FALSE}
summary(WhiteWineData$fixed.acidity)
```

```{r echo=FALSE}
hist1 <- qplot(data = WhiteWineData, x = WhiteWineData$fixed.acidity, binwidth = .1,
      xlab = "g / dm^3",
      color = I("black"),
      fill = I("pink"))+
  scale_x_continuous(lim = c(3,15), breaks = seq(3,15,1))

box1 <- qplot(x = quality, y = fixed.acidity, data = WhiteWineData, geom = 'boxplot',
              xlab = "rating",
              ylab = "g / dm^3")+
  coord_cartesian(ylim=c(5,9))

grid.arrange(hist1, box1)
```

### 2. Volatile Acidity (acetic acid)

"Volatile Acidity" was measured by the amount of acetic acid in the wine in grams per cubic decimeter.  The normal amount found in wine is around 0.3 g/dm^3.  This acid contributes to the smell and taste of the wine.  As the amount of acetic acid increases above 0.3 g/dm^3, the wine develops a sour, vinegar taste to it.  Acetic acid levels in this data set appear to be slightly positively skewed.  The range within the 1st quartile was 0.13 g/dm^3, while the range within the 4th quartile was 0.78 g/dm^3 with 50% of the data falling between 0.27 g/dm^3 and 0.39 g/dm^3

```{r echo=FALSE}
summary(WhiteWineData$volatile.acidity)
```

```{r echo=FALSE}
hist2 <- qplot(data = WhiteWineData, x = WhiteWineData$volatile.acidity, binwidth = .01,
      xlab = "g / dm^3",
      color = I("black"),
      fill = I("red"))+
  scale_x_continuous(lim = c(0,1.5), breaks = seq(0,1.5,.1))

box2 <- qplot(x = quality, y = volatile.acidity, data = WhiteWineData, geom = 'boxplot',
              xlab = "rating",
              ylab = "g / dm^3")+
  coord_cartesian(ylim=c(.1,.6))

grid.arrange(hist2, box2)
```

### 3. Citric Acid

The amount of citric acid in the wine was measured in grams per cubic decimeter.  This acid is used in small quantities to lower the pH (make it slightly more acidic) as well as add 'freshness' and flavor.  The distribution of this variable in the dataset is slightly positively skewed, with a mean of 0.3342 g/dm^3 and a median of 0.3200 g/dm^3.  This dataset contained outliers with a range of 1.27 g/dm^3 within the 4th quartile. 

```{r echo=FALSE}
summary(WhiteWineData$citric.acid)
```

```{r echo=FALSE}
hist3 <- qplot(data = WhiteWineData, x = WhiteWineData$citric.acid, binwidth = .05,
      xlab = "g / dm^3",
      color = I("black"),
      fill = I("orange"))+
  scale_x_continuous(lim = c(0,2), breaks = seq(0,2,.2))

box3 <- qplot(x = quality, y = citric.acid, data = WhiteWineData, geom = 'boxplot',
              xlab = "rating",
              ylab = "g / gm^3")+
  coord_cartesian(ylim=c(.1,.6))

grid.arrange(hist3, box3)
```

### 4. Residual Sugar

"Residual Sugar" is a measure of the amount of sugar remaining after fermentation stops in grams per cubic decimeter. It's rare to find wines with less than 1 g/dm^3 and wines with greater than 45 g/dm^3 are considered sweet.  The outliers in this dataset cause the distribution to be non-normal with the range within the 4th quartile of 55.9 g/dm^3 and 75% of the data falling below 9.9 g/dm^3.  After transforming this data to a log10 scale, it appears the data is bimodal.

```{r echo=FALSE}
summary(WhiteWineData$residual.sugar)
```

```{r echo=FALSE}
p1 <- qplot(data = WhiteWineData, x = WhiteWineData$residual.sugar,
      xlab = "g / dm^3",
      color = I("black"),
      fill = I("dark grey"))

p2 <- p1 + scale_x_log10()

p3 <- qplot(x = quality, y = residual.sugar, data = WhiteWineData, geom = 'boxplot',
            xlab = "rating",
            ylab = "g / dm^3")+
  coord_cartesian(ylim=c(0,15))

grid.arrange(p1, p2, p3)
```

### 5. Chlorides (sodium chloride)

"Chlorides" were the measure of the amount of sodium chloride (salt) in the wine in g/dm^3.  This dataset has a positively skewed distribution with 75% of the data falling below 0.05 g/dm^3 and 25% of the data falling between 0.05 g/dm^3 and 0.346 g/dm^3.  Looking at the boxplots of Chloride distribtion across each quality rating, it looks like there might be a trend of higher ratings for wines with lower salt content, which I will evaluate further in later sections.

```{r echo=FALSE}
summary(WhiteWineData$chlorides)
```

```{r echo=FALSE}
hist4 <- qplot(data = WhiteWineData, x = WhiteWineData$chlorides, binwidth = .005,
      xlab = "g / dm^3",
      color = I("black"),
      fill = I("blue"))+
  scale_x_continuous(lim = c(0,.35), breaks = seq(0,.35,.05))

box4 <- qplot(x = quality, y = chlorides, data = WhiteWineData, geom = 'boxplot',
              xlab = "rating",
              ylab = "g / dm^3")+
  coord_cartesian(ylim=c(0.01,.075))

grid.arrange(hist4, box4)
```

### 6. Free Sulfur Dioxide

Free Sulfur Dioxide was measured in mg/dm^3.  The free form of SO2 exists in equilibrium between molecular SO2 (as a dissolved gas) and bisulfite ion.  It prevents microbial growth and the oxidation of wine.  This variable in the dataset is positively skewed because of larger numbers within a range of 243 mg/dm^3 in the 4th quartile of this dataset.  The median is 34 mg/dm^3 and the mean is 35.31 mg/dm^3

```{r echo=FALSE}
summary(WhiteWineData$free.sulfur.dioxide)
```

```{r echo=FALSE}
hist5 <- qplot(data = WhiteWineData, x = WhiteWineData$free.sulfur.dioxide, binwidth = 5,
      xlab = "mg / dm^3",
      color = I("black"),
      fill = I("yellow"))+
  scale_x_continuous(lim = c(0,300), breaks = seq(0,300,20))

box5 <- qplot(x = quality, y = free.sulfur.dioxide, data = WhiteWineData, geom = 'boxplot',
              xlab = "rating",
              ylab = "mg / dm^3")+
  coord_cartesian(ylim=c(0,75))

grid.arrange(hist5, box5)
```

### 7. Total Sulfur Dioxide

"Total Sulfur Dioxide" is the amount of free and bound forms of S02.  In low concentrations, SO2 is mostly undetectable in wine, but at free SO2 concentrations over 50 ppm, SO2 becomes evident in the smell and taste of wine.  Outliers skew this distribution in the positive direction with a median of 134 mg/dm^3 and a mean of 138.4 mg/dm^3.  50% of the data falls between 108 mg/dm^3 and 167 mg/dm^3. 

```{r echo=FALSE}
summary(WhiteWineData$total.sulfur.dioxide)
```

```{r echo=FALSE}
hist6 <- qplot(data = WhiteWineData, x = WhiteWineData$total.sulfur.dioxide, binwidth = 10,
      xlab = "mg / dm^3",
      color = I("black"),
      fill = I("brown"))+
  scale_x_continuous(lim = c(0,450), breaks = seq(0,450,20))

box6 <- qplot(x = quality, y = total.sulfur.dioxide, data = WhiteWineData, geom = 'boxplot',
              xlab = "rating",
              ylab = "mg / dm^3")+
  coord_cartesian(ylim=c(50,250))

grid.arrange(hist6, box6)
```

### 8. Density

"Density" is determined by the alcohol and sugar content of the wine in grams per cubic centimeter.  50% of the data falls beween 0.9917 g/cm^3 and 0.9961 g/cm^3.  The median is 0.9937 and the mean is 0.9940.  This distribution in slightly positively skewed because of outliers falling between 0.9961 g/cm^3 and 1.039 g/cm^3.  The box plot distributions across categorical ratings show there might be a relationship between the density of wine and it's subsequent rating (The lower the density, the higher the rating), so this will be worth evalutating in later sections.

```{r echo=FALSE}
summary(WhiteWineData$density)
```

```{r echo=FALSE}
hist7 <- qplot(data = WhiteWineData, x = WhiteWineData$density, binwidth = .001,
      xlab = "g / cm^3",
      color = I("black"),
      fill = I("green"))+
  scale_x_continuous(lim = c(.98,1.05), breaks = seq(.98,1.05,.01))

box7 <- qplot(x = quality, y = density, data = WhiteWineData, geom = 'boxplot',
              xlab = "rating",
              ylab = "g / cm^3")+
  coord_cartesian(ylim=c(.985,1))

grid.arrange(hist7, box7)
```

### 9. pH

"pH" describes how acidic or basic a wine is on a scale from 0 (very acidic) to 14 (very basic); most wines are between 3-4 on the pH scale.  This variable is normally distributed in this dataset with a mean of 3.188 and a median of 3.180. 

```{r echo=FALSE}
summary(WhiteWineData$pH)
```

```{r echo=FALSE}
hist8 <- qplot(data = WhiteWineData, x = WhiteWineData$pH, binwidth = .05,
      xlab = "pH",
      color = I("black"),
      fill = I("purple"))+
  scale_x_continuous(lim = c(2.5,4), breaks = seq(2.5,4,.1))

box8 <- qplot(x = quality, y = pH, data = WhiteWineData, geom = 'boxplot',
              xlab = "rating",
              ylab = "pH")+
  coord_cartesian(ylim=c(2.9,3.5))

grid.arrange(hist8, box8)
```

### 10. Sulphates (postassium sulphate)

"Sulphates" are a measurement of the amount of postassium sulphate in g/dm^3.  Potassium sulphate is a wine additive which can contribute to sulfur dioxide gas (S02) levels, which acts as an antimicrobial and antioxidant.  This variable is slightly positively skewed, with a median of 0.47 g/dm^3 and a mean of 0.4898 g/dm^3.  50% of the data falls between 0.41 g/dm^3 and 0.55 g/dm^3. 

```{r echo=FALSE}
summary(WhiteWineData$sulphates)
```

```{r echo=FALSE}
hist9 <- qplot(data = WhiteWineData, x = WhiteWineData$sulphates, binwidth = .05,
      xlab = "g / dm^3",
      color = I("black"),
      fill = I("dark green"))+
  scale_x_continuous(lim = c(0,1.1), breaks = seq(0,1.1,.1))

box9 <- qplot(x = quality, y = sulphates, data = WhiteWineData, geom = 'boxplot',
              xlab = "rating",
              ylab = "g / dm^3")+
  coord_cartesian(ylim=c(.3,.65))

grid.arrange(hist9, box9)
```

### 11. Alcohol

"Alcohol" is a measurement of the percentage of alcohol content by volume found in the wine.  This variable is positively skewed with 75% of the data falling between 8% and 11.4%, while the remaining 25% of data fell between 11.4% and 14.2%.  The boxplots of alcohol content across each categorical rating shows a possible relationship between the quality of the wine and the alcohol content (the higher the alcohol content, the better the quality rating).  I will dig into this further in later sections.  

```{r echo=FALSE}
summary(WhiteWineData$alcohol)
```

```{r echo=FALSE}
hist10 <- qplot(data = WhiteWineData, x = WhiteWineData$alcohol, binwidth = .5,
      xlab = "% by volume",
      color = I("black"),
      fill = I("light blue"))+
  scale_x_continuous(lim = c(5,15), breaks = seq(5,15,1))

box10 <- qplot(x = quality, y = alcohol, data = WhiteWineData, geom = 'boxplot',
              xlab = "rating",
              ylab = "% by volume")+
  coord_cartesian(ylim=c(9,13))

grid.arrange(hist10, box10)
```

### 12. Quality

The quality of the wine was measured as the average quality of each wine, which was rated from 0-10 (0 = very bad, 10 = very excellent) by at least 3 experts based on sensory data.  This distribution would ideally have the same amount of observations for each categorical rating, however there are no observations of wines with ratings below 2 or any with a perfect 10 rating.  92.6% of the wines in this dataset were rated between 5-7.  This is a significant disadvantage for being able to draw conclusions on the differences between very excellent and very bad wine.

```{r echo=FALSE}
summary(WhiteWineData$quality)
```

```{r echo=FALSE}
qplot(data = WhiteWineData, x = WhiteWineData$quality, 
      xlab = "rating",
      color = I("black"),
      fill = I("light grey"))

```

# Univariate Analysis

### What is the structure of your dataset?

This dataset includes 4,898 observations of wine with 11 numerical input variables of chemical properties (fixed acidity, volatile acidity, citric acid, residual sugar, chlorides, free sulfer dioxide, total sulfur dioxide, density, pH, sulphates and alcohol) and 1 output variable of quality rating on a scale of 0-10.  

```{r echo=FALSE}
str(WhiteWineData)
```

### What is/are the main feature(s) of interest in your dataset?

The guiding question of this dataset is "Which chemical properties influence the quality of white wines?", so the main feature of interest is the quality rating.

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?

I think Chlorides, Density and Alcohol levels might show a relationship to the quality rating of each wine.  

### Did you create any new variables from existing variables in the dataset?

I did not create any new variables for this dataset.

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?

The residual sugar distribution was positively skewed by outliers, so tranforming the x-axis to a log10 scale uncovered a bimodal distribution among the majority of the data.  As well, I transformed the quality variable to an ordered factor so that box plots would show each individual rating as its own category.  

# Bivariate Plots Section

> **Tip**: Based on what you saw in the univariate plots, what relationships
between variables might be interesting to look at in this section? Don't limit
yourself to relationships between a main output feature and one of the
supporting variables. Try to look at relationships between supporting variables
as well.

```{r echo=FALSE, Bivariate_Plots}

```

# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

### What was the strongest relationship you found?


# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!