---
title: Exploratory Data Analysis - White Wine Quality
author: Julia Fuhrman
output:
  html_document: default
  pdf_document: default
---
================================================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(ggthemes)
library(gridExtra)
library(dplyr)
library(GGally)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
setwd('/Users/julia/Documents/UdacityProjects/Data Analyst Nanodegree/Exploratory Data Analysis/final-project')
WhiteWineData <- read.csv('wineQualityWhites.csv')
theme_set(theme_minimal(12)) 
```

This analysis explores 11 chemical properties of white wine for 4,898 wine 
samples, as well as a rating of the quality of the wine on a 0-10 scale.

# Univariate Plots Section

### 1. Fixed Acidity (tartaric acid)

"Fixed Acidity" was measured by the amount of tartaric acid in the wine in grams
per cubic decimeter.  The distribution of this variable in the dataset is
normal, with a mean of 6.855 g/dm^3 and a median of 6.8 g/dm^3.

```{r echo=FALSE, Fixed_Acidity_Summary}
summary(WhiteWineData$fixed.acidity)
```

```{r echo=FALSE, Fixed_Acidity}
qplot(data = WhiteWineData,
      x = WhiteWineData$fixed.acidity, 
      binwidth = .1,
      xlab = "Fixed Acidity - g / dm^3",
      color = I("black"),
      fill = I("pink")) +
  scale_x_continuous(lim = c(3,15), 
                     breaks = seq(3,15,1))

```

### 2. Volatile Acidity (acetic acid)

"Volatile Acidity" was measured by the amount of acetic acid in the wine in 
grams per cubic decimeter.  The overall distribution of this variable is 
slightly positively skewed.  The range within the 1st quartile was 0.13 g/dm^3,
while the range within the 4th quartile was 0.78 g/dm^3 with 50% of the data 
falling between 0.27 g/dm^3 and 0.39 g/dm^3

```{r echo=FALSE, Volatile_Acidity_Summary}
summary(WhiteWineData$volatile.acidity)
```

```{r echo=FALSE, Volatile_Acidity}
qplot(data = WhiteWineData, 
      x = WhiteWineData$volatile.acidity, 
          binwidth = .01,
          xlab = "Volatile Acidity - g / dm^3",
          color = I("black"),
          fill = I("red")) +
  scale_x_continuous(lim = c(0,1.5), 
                     breaks = seq(0,1.5,.1))
```

### 3. Citric Acid

The amount of citric acid in the wine was measured in grams per cubic decimeter.
The distribution of this variable in the dataset is slightly positively skewed, 
with a mean of 0.3342 g/dm^3 and a median of 0.3200 g/dm^3.

```{r echo=FALSE, Citric_Acid_Summary}
summary(WhiteWineData$citric.acid)
```

```{r echo=FALSE, Citric_Acid}
qplot(data = WhiteWineData, 
      x = WhiteWineData$citric.acid, 
      binwidth = .05,
      xlab = "Citric Acid - g / dm^3",
      color = I("black"),
      fill = I("orange")) +
  scale_x_continuous(lim = c(0,2), breaks = seq(0,2,.2))
```

### 4. Residual Sugar

"Residual Sugar" is a measure of the amount of sugar remaining after 
fermentation stops in grams per cubic decimeter. It's rare to find wines with 
less than 1 g/dm^3 and wines with greater than 45 g/dm^3 are considered sweet. 
The outliers in this dataset cause the distribution to be non-normal with the 
range within the 4th quartile of 55.9 g/dm^3 and 75% of the data falling below 
9.9 g/dm^3.  After transforming this data to a log10 scale, it appears the data 
is bimodal. 

```{r echo=FALSE, Residual_Sugar_Summary}
summary(WhiteWineData$residual.sugar)
```

```{r echo=FALSE, Residual_Sugar}
residual.sugar1 <- qplot(data = WhiteWineData, 
                         x = WhiteWineData$residual.sugar,
                         binwidth = 1,
                         xlab = "Residual Sugar - g / dm^3",
                         color = I("black"),
                         fill = I("dark grey"))

residual.sugar2 <- qplot(data = WhiteWineData, 
                         x = WhiteWineData$residual.sugar,
                         xlab = "Residual Sugar - g / dm^3",
                         binwidth = .1,
                         color = I("black"),
                         fill = I("dark grey")) + 
  scale_x_log10(breaks = seq(1,10,1))

grid.arrange(residual.sugar1, residual.sugar2)
```

### 5. Chlorides (sodium chloride)

"Chlorides" were the measure of the amount of sodium chloride (salt) in the wine
in g/dm^3.  This dataset has a positively skewed distribution with 75% of the
data falling below 0.05 g/dm^3 and 25% of the data falling between 0.05 g/dm^3 
and 0.346 g/dm^3. 

```{r echo=FALSE, Chlorides_Summary}
summary(WhiteWineData$chlorides)
```

```{r echo=FALSE, Chlorides}
qplot(data = WhiteWineData, 
      x = WhiteWineData$chlorides, 
      binwidth = .005,
      xlab = "Chlorides - g / dm^3",
      color = I("black"),
      fill = I("blue")) +
  scale_x_continuous(lim = c(0,.35), breaks = seq(0,.35,.05))
```

### 6. Free Sulfur Dioxide

Free Sulfur Dioxide was measured in mg/dm^3.  The free form of SO2 exists in 
equilibrium between molecular SO2 (as a dissolved gas) and bisulfite ion.  It 
prevents microbial growth and the oxidation of wine.  This variable in the 
dataset is positively skewed because of larger numbers within a range of 
243 mg/dm^3 in the 4th quartile of this dataset.  The median is 34 mg/dm^3 and 
the mean is 35.31 mg/dm^3.  

```{r echo=FALSE, Free_Sulfur_Dioxide_Summary}
summary(WhiteWineData$free.sulfur.dioxide)
```

```{r echo=FALSE, Free_Sulfur_Dioxide}
qplot(data = WhiteWineData, 
      x = WhiteWineData$free.sulfur.dioxide, 
      binwidth = 5,
      xlab = "Free Sulfur Dioxide - mg / dm^3",
      color = I("black"),
      fill = I("yellow")) +
  scale_x_continuous(lim = c(0,300), breaks = seq(0,300,20))

```

### 7. Total Sulfur Dioxide

"Total Sulfur Dioxide" is the amount of free and bound forms of S02.  In low 
concentrations, SO2 is mostly undetectable in wine, but at free SO2 
concentrations over 50 ppm, SO2 becomes evident in the smell and taste of wine.
Outliers skew this distribution in the positive direction with a median of 
134 mg/dm^3 and a mean of 138.4 mg/dm^3.  50% of the data falls between 
108 mg/dm^3 and 167 mg/dm^3.

```{r echo=FALSE, Total_Sulfur_Dioxide_Summary}
summary(WhiteWineData$total.sulfur.dioxide)
```

```{r echo=FALSE, Total_Sulfur_Dioxide}
qplot(data = WhiteWineData,
      x = WhiteWineData$total.sulfur.dioxide, 
      binwidth = 10,
      xlab = "Total Sulfur Dioxide - mg / dm^3",
      color = I("black"),
      fill = I("brown")) +
  scale_x_continuous(lim = c(0,450), breaks = seq(0,450,50))
```

### 8. Density

"Density" is determined by the alcohol and sugar content of the wine in grams 
per cubic centimeter.  50% of the data falls beween 0.9917 g/cm^3 and 0.9961 
g/cm^3.  The median is 0.9937 and the mean is 0.9940.  This distribution in 
slightly positively skewed because of outliers falling between 0.9961 g/cm^3 and
1.039 g/cm^3.

```{r echo=FALSE, Density_Summary}
summary(WhiteWineData$density)
```

```{r echo=FALSE, Density}
qplot(data = WhiteWineData, 
      x = WhiteWineData$density,
      binwidth = .001,
      xlab = "Density - g / cm^3",
      color = I("black"),
      fill = I("green")) +
  scale_x_continuous(lim = c(.98,1.05), 
                     breaks = seq(.98,1.05,.01))
```

### 9. pH

"pH" describes how acidic or basic a wine is on a scale from 0 (very acidic) to 
14 (very basic); most wines are between 3-4 on the pH scale.  This variable is 
normally distributed in this dataset with a mean of 3.188 and a median of 3.180.

```{r echo=FALSE, pH_Summary}
summary(WhiteWineData$pH)
```

```{r echo=FALSE, pH}
qplot(data = WhiteWineData, 
      x = WhiteWineData$pH, 
      binwidth = .05,
      xlab = "pH",
      color = I("black"),
      fill = I("purple"))+
  scale_x_continuous(lim = c(2.5,4), 
                     breaks = seq(2.5,4,.1))
```

### 10. Sulphates (postassium sulphate)

"Sulphates" are a measurement of the amount of postassium sulphate in g/dm^3.  
Potassium sulphate is a wine additive which can contribute to sulfur dioxide gas
(S02) levels, which acts as an antimicrobial and antioxidant.  This variable is 
slightly positively skewed, with a median of 0.47 g/dm^3 and a mean of 0.4898 
g/dm^3.  50% of the data falls between 0.41 g/dm^3 and 0.55 g/dm^3.

```{r echo=FALSE, Sulphates_Summary}
summary(WhiteWineData$sulphates)
```

```{r echo=FALSE, Sulphates}
qplot(data = WhiteWineData,
      x = WhiteWineData$sulphates, 
      binwidth = .05,
      xlab = "Sulphates - g / dm^3",
      color = I("black"),
      fill = I("dark green"))+
  scale_x_continuous(lim = c(0,1.1),
                     breaks = seq(0,1.1,.1))

```

### 11. Alcohol

"Alcohol" is a measurement of the percentage of alcohol content by volume found 
in the wine.  This variable is positively skewed with 75% of the data falling 
between 8% and 11.4%, while the remaining 25% of data fell between 11.4% and 
14.2%.

```{r echo=FALSE, Alcohol_Summary}
summary(WhiteWineData$alcohol)
```

```{r echo=FALSE, Alcohol}
qplot(data = WhiteWineData, 
      x = WhiteWineData$alcohol, 
      binwidth = .5,
      xlab = "Alcohol - % by volume",
      color = I("black"),
      fill = I("light blue"))+
  scale_x_continuous(lim = c(5,15), 
                     breaks = seq(5,15,1))
```

### 12. Quality

Quality was measured by taking the average of 3 expert ratings on the quality of
the wine, which was rated on a 10 point scale, 0 being "really bad" and 10 being
"really excellent".  This dataset had a disproportionate sampling of each 
rating, with no samples for 0, 1, 2, or 10 rated wines and very few for 3, 4, 8,
and 9 rated wines.  74.6% of the samples were rated as a 5 or 6.  

```{r echo=FALSE, Quality_Summary}
table(WhiteWineData$quality)
```

```{r echo=FALSE, Quality}
qplot(data = WhiteWineData, 
      x = WhiteWineData$quality, 
      binwidth = 1,
      xlab = "Quality Rating",
      color = I("black"),
      fill = I("yellow green")) +
  scale_x_continuous(breaks=seq(3,9,1))
```

# Univariate Analysis

### What is the structure of your dataset?

This dataset includes 4,898 observations of wine with 11 numerical input
variables of chemical properties (fixed acidity, volatile acidity, citric acid, 
residual sugar, chlorides, free sulfer dioxide, total sulfur dioxide, density, 
pH, sulphates and alcohol) and 1 output variable of quality rating on a scale of
0-10, which I transformed into a new variable "quality.bins" to group the 
quality into 3 factor levels.  

```{r echo=FALSE}
str(WhiteWineData)
```

### What is/are the main feature(s) of interest in your dataset?

The guiding question of this dataset is "Which chemical properties influence the
quality of white wines?", so the main feature of interest is the quality rating.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

I think Chlorides, Density, Alcohol, Volatile Acidity and Free Sulphur Dioxide
levels might show a relationship to the quality rating of each wine.  Fixed 
Acidity and pH might also contribute slightly to the quality ratings.

### Did you create any new variables from existing variables in the dataset?

Yes, I decided to create a new variable called "quality.buckets" to 
visualize the correlations found using Spearman's method in the bivariate
analysis below.  

I also created the variable "sugar.alcohol.ratio" during my multivariate 
analysis to see how this relationship affected the density and quality ratings.


### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

The residual sugar distribution was positively skewed by outliers, so 
tranforming the x-axis to a log10 scale uncovered a bimodal distribution among 
the majority of the data. 

# Bivariate Plots Section

### Comparing Quality 

Because quality is an ordinal variable and the main feature of interest, I used
Spearmans method to calculate the correlation between quality and each other 
variable in the dataset.  The highest correlated variables with quality were
alcohol, density and chlorides.  Using the 0-10 rating scale to visualize the
distribution of these variables made it difficult to see any patterns, so I 
grouped the ratings into 3 buckets: "bad", "average" and "good" wines.  This
revealed a clearer pattern where "good" quality ratings correlated with higher
alcohol content, lower density and lower chlorides compared to "bad" and 
"average" wines.

```{r echo=FALSE, Box_Plots}

# Create new quality buckets
WhiteWineData$quality.buckets <- cut(WhiteWineData$quality, 
                                  breaks = c(0, 4, 6, 10), 
                                  labels = c('bad','average','good'))

# Create boxplots with mean markers
alcohol_by_quality <- ggplot(aes(x = quality.buckets, 
                                 y = alcohol),
                             data = WhiteWineData) + 
  geom_boxplot() +
  stat_summary(fun.y = mean, 
               geom = 'point', 
               shape = 4) +
  coord_cartesian(ylim=c(9,13))

density_by_quality <-ggplot(aes(x = quality.buckets, 
                                y = density),
                            data = WhiteWineData) + 
  geom_boxplot() +
  stat_summary(fun.y = mean, 
               geom = 'point', 
               shape = 4) +
  coord_cartesian(ylim=c(.9875,.9975))
  
chlorides_by_quality <-ggplot(aes(x = quality.buckets, 
                                  y = chlorides),
                              data = WhiteWineData) + 
  geom_boxplot() +
  stat_summary(fun.y = mean, 
               geom = 'point', 
               shape = 4) +
  coord_cartesian(ylim=c(0.02,.06))
  
grid.arrange(alcohol_by_quality, density_by_quality, chlorides_by_quality)
```

### Correlation Matrix of Remaining Features

I was also interested to see how the chemical property variables correlated with
eachother to see if there were any relationships that strengthed the 
relationship with quality.  The top correlations (using pearsons method)
were density vs residual sugar and density vs alcohol.  There was also a 
moderate correlation between free sulfur dioxide vs total sulfur dioxide
(0.616), but I did not investigate this further since it's obvious that there 
would be a relationship since free sulfur dioxide is a component of the total
sulfur dioxide.

```{r echo=FALSE, Scatterplot_Matrix}
ggcorr(WhiteWineData[,2:12], 
       label = TRUE, 
       label_size = 3,
       label_round = 2,
       hjust = 0.8, 
       size = 3,
       layout.exp = 1)
```

### Density vs Residual Sugar

The relationship between density and residual sugar had the strongest 
correlation between any 2 variables (0.84).  Looking at the plot, it appears 
that this relationship is non-linear.  As density increases up to around
0.995 g / cm^3 residual sugar stays relatively the same, but increases at a
faster rate after density levels of 0.995 g / cm^3

```{r echo=FALSE, Density_vs_Residual_Sugar}
ggplot(aes(x = density, 
           y = residual.sugar), 
       data=WhiteWineData) +
  geom_jitter(alpha = 0.1) +
  coord_cartesian(xlim=c(0.985,1.01),
                  ylim=c(0,30)) +
  geom_smooth(method='auto', linetype = 2) +
  geom_vline(xintercept=0.995, color = 'red')
```

### Density vs Alcohol

The second strongest correlation was density vs alcohol (-0.78).  This also 
appears to be a non-linear relationship.  Alcohol levels decrease rapidly as 
density levels approach 0.995 g / cm^3 and then alcohol levels decrease at a 
much slower rate and almost level out after the 0.995 g / cm^3 threshold.  

```{r echo=FALSE, Density_vs_Alcohol}
ggplot(aes(x = density, 
           y = alcohol), 
       data=WhiteWineData) +
  geom_jitter(alpha = 0.15) +
  coord_cartesian(xlim=c(0.985,1.01)) +
  geom_smooth(method='auto', linetype = 2) +
  geom_vline(xintercept=0.995, color = 'red')
```


# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

Since "quality" is ordinal, I used Spearmans correlation to compare each 
variable to the quality ratings.  The highest correlations were:

```{r echo=FALSE, warning = FALSE, Quality_Comparisons}
with(WhiteWineData, cor.test(quality, chlorides, method='spearman'))

with(WhiteWineData, cor.test(quality, density, method='spearman'))

with(WhiteWineData, cor.test(quality, alcohol, method='spearman'))
```

As quality increases, alcohol content increases, while density and chlorides
decrease

### What was the strongest relationship you found?
The strongest relationship I found to the feature of interest (quality) was 
alcohol.  Interestingly, when looking at the wine in my house, the only
component on the labels that I could use for comparison when shopping for wine
is the alcohol % by volume.  Looking at the mean and median alcohol content for
each quality rating, a good rule of thumb is a good quality wine will have 
approximately 11% alcohol by volume or more.

```{r echo=FALSE, Alcohol_Quality_Summary}
quality_groups <- group_by(WhiteWineData, quality)
alcohol_by_quality <- summarise(quality_groups,
                          alcohol_mean = mean(alcohol),
                          alcohol_median = median(alcohol),
                          n = n())
alcohol_by_quality <- arrange(alcohol_by_quality, quality)
alcohol_by_quality
```

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

The top 2 strongest relationships I found among the remaining features were:

- Density and Residual Sugar = 0.839
- Density and Alcohol = -0.78

Looking at the plots, it looks like at around 0.993 Density, residual sugar 
increases slowly, and then after 0.993 the rate of increase is larger.  
Alcohol content on the other hand decreases at a high rate and then tapers off
around this same density threshold.  

# Multivariate Plots Section

### Alcohol, Density, Quality

Alcohol had the strongest correlation to quality, and density had the strongest
correlation to alcohol, so I was curious to see how quality was affected by the
relationship between density and alcohol.  The plot below shows that "good" 
quality wines are clustered around lower density and higher alcohol levels.  
This also reinforces the finding in the bivariate plot of alcohol vs quality 
where alcohol levels above 11 and density levels below 0.995 correspond to 
higher quality wines.

```{r echo=FALSE, Quality_Alcohol_Density}
ggplot(aes(x=alcohol,
           y = density), 
       data = WhiteWineData) +
  geom_point(aes(color=factor(quality.buckets))) +
  coord_cartesian(ylim=c(0.985, 1.005)) +
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality')) +
  geom_vline(xintercept = 11, color = 'red', linetype = 2) +
  geom_hline(yintercept = 0.995, color = 'red', linetype = 2)
```

### Alcohol, Sugar, Density

Since the strongest correlations in the whole dataset were density vs residual
sugar and density vs alcohol, I was curious to see how these 3 variables 
interacted with eachother.  It is interesting that at density levels below 0.995
g / cm^3, residual sugar levels appear in distinct lines around 1 1 g/dm^3, and
then appear much more scattered at higher levels.  It looks like the best ratio
of sugar to alcohol to achieve a low density wine is low residual sugar levels
to high alcohol levels.

```{r echo=FALSE, Alcohol_Density_Sugar}
ggplot(aes(x=residual.sugar,
           y = density), 
       data = WhiteWineData) +
  geom_point(aes(color=alcohol)) +
  scale_x_log10() +
  coord_cartesian(ylim=c(0.985, 1.005)) +
  geom_smooth(method='auto', color = 'red', linetype = 2)
```

### Alcohol, Sugar, Density, Quality

To confirm my finding that the best ratio to acheive a low density is low sugar
to high alcohol, I wanted to see if this maintained the relationship between
density and quality (where higher quality wines had lower densities).  I created
a new variable to quantify the sugar to alcohol ratio.  The plot below shows
that indeed, ratios below 1 (low sugar, high alcohol) had a bigger cluster of
"good" quality wines at low density.  

```{r echo=FALSE, Density_vs_Sugar_Alcohol_Ratio}

#Create sugar/alcohol ratio variable
WhiteWineData <- transform(WhiteWineData,
                           sugar.alcohol.ratio = residual.sugar/alcohol)

ggplot(aes(x=sugar.alcohol.ratio, 
           y = density), 
       data = WhiteWineData) +
  geom_point(aes(color =factor(quality.buckets))) +
  coord_cartesian(ylim=c(.985,1.005),
                  xlim=c(0,2)) +
  geom_vline(xintercept=1) +
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality')) +
  geom_smooth(method='lm', color = 'red', linetype = 2)
```

### Sugar, Alcohol, Quality

I made one final visualization, removing density from the analysis to see the 
relationship between sugar, alcohol and quality.  I found it interesting that
just like the residual sugar vs density plot above, alcohol levels appear in
distinct lines at sugar levels around 1 g/dm^3 and then appear more scattered at
higher levels.  This plot shows that "good" quality wines are clusted at the 
highest alcohol levels, regardless of sugar levels. 

```{r echo=FALSE, sugar_vs_alcohol_by_quality}
ggplot(aes(x=residual.sugar,
           y = alcohol), 
       data = WhiteWineData) +
  geom_point(aes(color=factor(quality.buckets))) +
  scale_x_log10() +
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality')) +
  geom_hline(yintercept = 11, color = 'red', linetype = 2)
```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest? Were there any interesting or \
surprising interactions between features?

It was clear in the first plot that higher quality wines have a low density with
a high alcohol % by volume.  As well, this relationship is maintained with a 
low residual sugar content.  Overall, quality wines have a smaller sugar to
alcohol ratio for a lower density level.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
ggplot(aes(x = quality.buckets, 
           y = alcohol),
       data = WhiteWineData) + 
  geom_boxplot() +
  stat_summary(fun.y = mean, 
               geom = 'point', 
               shape = 4) +
  coord_cartesian(ylim=c(9,13)) +
  labs(title = "Alcohol by Quality") +
  ylab("Alcohol - % by volume") +
  xlab("Quality")
```

### Description One
The highest correlation with the quality ratings of white wine was the alcohol
% by volume.  Higher quality wines have more than 11% alcohol by volume.

### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(aes(x=alcohol,
           y = density), 
       data = WhiteWineData) +
  geom_point(aes(color=factor(quality.buckets))) +
  coord_cartesian(ylim=c(0.985, 1.005)) +
  labs(title = "Alcohol Content vs Density by Quality",
       color = "Quality") +
  ylab("Density - g/cm^3") +
  xlab("Alcohol - % by volume") +
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality')) +
  geom_vline(xintercept = 11.5, color = 'red', linetype = 2) +
  geom_hline(yintercept = 0.995, color = 'red', linetype = 2)
```

### Description Two
Density was also correlated with quality.  Higher quality wines had a lower 
density compared to bad and average wines.  When compared in combination to 
alcohol levels, the relationships are maintained.  Quality wines have a lower
density (below approx. 0.995) with a higher alcohol content (ideally above
11.5%).

### Plot Three
```{r echo=FALSE, Plot_Three}
ggplot(aes(x=residual.sugar,
           y = density), 
       data = WhiteWineData) +
  geom_point(aes(color=alcohol)) +
  scale_x_log10() +
  coord_cartesian(ylim=c(0.985, 1.005)) +
  labs(title = "Residual Sugar vs Density by Alcohol Content") +
  ylab("Density - g/cm^3") +
  xlab("Residual Sugar - g/dm^3") +
  geom_smooth(method='auto', color = 'red', linetype = 2)
```

### Description Three
The highest correlation was between residual sugar and density.  Adding in a 
comparison with alcohol content, it is clear that the optimal ratio for a low
density wine is lower residual sugar content to higher alcohol content.

------

# Reflection

The white wine dataset contains 11 chemical properties of 4,898 samples of wine,
as well as an average quality rating by 3 experts.  I started by looking at the
distributions of each chemical component and then compared each component to the
quality ratings.  The residual sugar distribution had to be transformed to a 
log10 scale to be able to see a closer to normal distribution, which 
interestingly uncovered a bimodal distribution. All other distributions were
normally distributed so I was able to use pearsons method to calculate 
correlations between chemical components and spearmans method to calculate 
correlations to the ordinal "quality" variable.  As well, I compared chemical
properties to eachother and visualized how those relationship affected the 
quality ratings.  

There was a correlation between alcohol % by volume to the quality of wine as
well as a correlation between the density of the wine to the quality rating. The 
highest correlation between chemical components was density vs residual sugar.  
In comparing alcohol, residual sugar, density and quality in various 
multivariate plots, I found that the ideal ratio for a low density and therefore
high quality wine, was lower residual sugar levels to higher alcohol levels.

The biggest limitation of this dataset is the disproportionate representation of
each quality level rating.  There were no samples of 0, 1, 2 or 10 rated wines.
Additionally, there were significantly fewer examples of wines rated at a 3, 4,
8 and 9.  The majority of the data was for average rated wines, which limits the
ability to see trends between bad, average, good, and excellent wines.  By 
creating 3 levels ("bad", "average", and "good") I was better able to visualize
difference among each category.  Another limitation is that this dataset is from 
variants of the Portuguese "Vinho Verde" wine.  Further interesting analysis 
would include comparisons of different regions/countries, types of grapes used 
and prices these wines are sold at.
